Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-10-31 manuscript.tex
latest

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig. 4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig. \ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.

@@

The importance of $b$ at all densities is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript. The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition. In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from

>>>>

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig. 4) refers instead to the direction of trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Yet Fig. \ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas improvements in $c$ yield an appreciable fitness benefit.

@@

The importance of $b$ at all densities is a general feature of lottery competition. Indeed, as can be seen from Eq. \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}. This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists. This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}. It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript. The point of generalizing the classic lottery model here is to account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition. There is nothing preventing our model from also being applied to study species invasion, but our focus here is on genotypes within a population rather than the interactions between different species.

Given
